---
title: "TimeSeries_Analysis of Fruit-Vege(2012-2021)"
author: "Mecailla Lomoljo"
date: "28 November 2022"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


```

## Fruit Vegetables: Retail Prices of Agricultural Commodities by Geolocation, Commodity, Year and Period


This data is a real data from Philippine Statistics Authority Openstat website: <https://openstat.psa.gov.ph/>. In this notebook we will perform a Time series Analysis on the Fruit Vegetables: Retail Prices of Agricultural Commodities by Geolocation, Commodity, Year and Period. This data is limited to Ncr only, and a 2012-2021 based data.

```{r loading packages, include=FALSE}

library(tidyverse)
library(dplyr)
library(zoo)
library(forecast)
library(dlookr)
library(TSstudio)
library(ggplot2)
library(readxl)
library(lubridate)
library(xts)
library(Hmisc)
library(rcompanion)

```
```{r import file}

fruit_vege <- read_excel("C:/Users/Mecailla/Documents/TimeSeries/fruit_vege.xlsx")
head(fruit_vege)

```


### Data Exploration

Looking at the shape of our Data set, it contains 120 rows/observation and 15 columns/variables.
These variables are:
* ampalaya, sayote, upo, pipino, talong, patola, squash, tomato, and events;
* Pandemic, Habagat, Amihan, Volcanic Eruption, El Nino, La nina

And Date as our index.

Now let's investigate our Data further.

```{r Structure}

head(fruit_vege)
colnames(fruit_vege)
str(fruit_vege)
```
I I extract month and year from Date to easily calculate Mean of every Fruit and Vegetables later.
```{r plotting1}

fruit_vege <- fruit_vege %>% mutate("Month"= format(fruit_vege$Date,"%m"), 
                                    "Year"= format(fruit_vege$Date,"%Y"))

```

Now, to look plot our mean prices per year.

```{r plot}

mean_per_year <- fruit_vege %>% group_by(Year) %>% summarise(Ampalaya= mean(ampalaya_1kg), Sayote= mean(sayote_1kg), Upo= mean(upo_1kg), Pipino= mean(pipino_1kg), Talong= mean(talong_1kg), Patola= mean(patola_1kg), Squash= mean(squash_1kg), Tomato= mean(tomato_1kg))

mean_per_year

#plotting the mean
mean_year_plot <- ts(mean_per_year, start= c(2012), end= c(2021))
plot(mean_year_plot, main = "Mean of Fruit Vegetables per Year",
     col= "blue", type= "b", lwd= 3)
```

I also wanted to see the plot of every fruit and vegetables through the years.

```{r}
frvg <- fruit_vege %>% select(ends_with("_1kg"))
ts_frvg <- ts(frvg, start= c(2012, 1), end= c(2021, 12), frequency = 12)

autoplot(ts_frvg, lwd = 1, alpha= 0.7)+
  labs(y="Price", x= "Date", title= "Prices of Fruit Vegetables vs Date") + 
  scale_y_log10()

```

If we check the data of our converted ts series data.


```{r observed}

start(ts_frvg)
end(ts_frvg)
cycle(ts_frvg)

```
 
##Decomposition
 
Since we have 8 variables to analyze, let me just
choose one variable for the sake of showing decomposition in R.

```{r subsetting}
ampalaya <- fruit_vege %>% select(ampalaya_1kg, pandemic:Year)
ampalaya_1 <- fruit_vege %>% select(ampalaya_1kg)
ts_ampalaya <- ts(ampalaya_1, start= c(2012, 1), end= c(2021, 12), frequency = 12)

```
```{r decomposing}
decomposed_ampalaya <- decompose(ts_ampalaya, type = "multiplicative")
autoplot(decomposed_ampalaya) + labs(main="Decomposition of Ampalaya")
autoplot(decomposed_ampalaya$random, main= "Residuals of Ampalaya")
plot(decomposed_ampalaya$seasonal, main= "Seasonal graph of Ampalaya")

summary(ts_ampalaya)

```

Checking for Normality of Residuals of Ampalaya

```{r}
qqnorm(decomposed_ampalaya$random)
qqline(decomposed_ampalaya$random)

shapiro.test(decomposed_ampalaya$random)
```

We run a multiple linear regression to our Ampalaya with Events as the predictor. 

```{r}
model_a <- lm(ampalaya_1kg~factor(pandemic)+factor(habagat)+
                factor(volcan_eruption) + factor(amihan)+ factor(Elnino) + factor(Lanina), 
              data=ampalaya )

model_b <- lm(ampalaya_1kg~factor(pandemic)+factor(habagat)+
                factor(volcan_eruption) + factor(amihan)+ factor(Elnino), 
              data=ampalaya )

model_c <- lm(ampalaya_1kg~factor(pandemic)+factor(habagat)+
                factor(volcan_eruption) + factor(amihan), 
              data=ampalaya )

model_d <- lm(ampalaya_1kg~factor(pandemic)+factor(habagat)+
                factor(volcan_eruption), 
              data=ampalaya )

model_e <- lm(ampalaya_1kg~factor(pandemic)+factor(habagat), 
              data=ampalaya )

model_f <- lm(ampalaya_1kg~factor(pandemic), 
              data=ampalaya )

model_g <- lm(ampalaya_1kg~+factor(habagat)+
                factor(volcan_eruption) + factor(Elnino), 
              data=ampalaya )


compareLM(model_a, model_b, model_c, model_d, model_e, model_f, model_g)
```

To check the correlation of our events to the price of Ampalaya, we run the following codes below:

```{r}
amp <- ampalaya %>% select(-c(Month, Year))
rcorr(as.matrix(amp))

```




